---
draft: true
output: html_document
editor_options:
  chunk_output_type: console
---


```{r}
#| include: false
#| eval: true

library(tidyverse)
library(tidycensus)
library(tigris)
library(r2d3)
library(leaflet)
library(sf)

library(dotenv)
library(here)

here::i_am('posts/Cartogram/shapefile.qmd')

load_dot_env(file = ".env")

census_api_key(Sys.getenv('CENSUS_KEY'))

options(tigris_use_cache = TRUE)

```

In my previous post, I used a cartogram as my visualization.

This post is a technical walkthrough of challenges and final implementation for a cartogram in Quarto.




```{r}

poverty_cd_s1701 <- get_acs(geography = "congressional district", 
        table = 'S1701',
        year = 2023,
        geometry = TRUE,
        cache_table = TRUE)

poverty_cd_s1701 %>% 
  glimpse()

v2023 <- load_variables(2023, "acs5/subject", cache = TRUE)

v2023 %>% glimpse()

v2023 %>% 
  filter(name == 'S1701_C01_001') %>% 
  glimpse()

data("fips_codes")

states <- fips_codes %>% 
  select(state, state_code) %>% 
  distinct()

poverty_cd_s1701_labels <- poverty_cd_s1701 %>% 
  left_join(v2023,
            by = c("variable" = "name")) %>% 
  mutate(state_code = substr(GEOID, 1, 2)) %>% 
  left_join(states)

poverty_cd_s1701_labels %>% 
  glimpse()
```

# Approaches

# geom_sf

```{r}


v2023 %>% 
  filter(name == 'S1701_C03_001') %>% 
  glimpse()


poverty_cd_s1701_labels %>% 
  filter(variable == 'S1701_C03_001') %>% 
  shift_geometry() %>% 
  ggplot(aes(fill = estimate)) + 
  geom_sf()

```


```{r}

poverty_cd_s1701_labels %>% 
  filter(variable == 'S1701_C03_001',
         state != 'PR') %>% 
  shift_geometry() %>% 
  ggplot(aes(fill = estimate)) + 
  geom_sf() + 
  scale_fill_distiller(palette = "RdPu", 
                       direction = 1) + 
  labs(title = "Poverty Rate By Congressional District",
       caption = "Data source: 2023 5-year ACS, US Census Bureau",
       fill = "ACS estimate") + 
  theme_void()

```


```{r}
#"C:\Users\mkuhn\Documents\projects\personal\DataBlog\posts\MedicaidFundingImpact\data\original\HexCDv31\HexCDv31.shp"
# library(geojsonsf)
library(geojsonio)


pov_rate_cd <-
  poverty_cd_s1701_labels %>% 
    filter(variable == 'S1701_C03_001',
           state != 'PR') %>%
    st_drop_geometry() %>% 
  tibble()

cd_hex_path = here('posts', 'Cartogram', 'data', 'original', 'HexCDv31', 'HexCDv31.shp')

cd_hex <-
  sf::read_sf(cd_hex_path)

cd_hex_pov <-
  cd_hex %>% 
    left_join(pov_rate_cd,
              by = 'GEOID')

# 
# cd_hex_geo <- geojson_json(cd_hex_pov) 
# 
# write_file(cd_hex_geo, 'cd_hex_geo_pov.json')

```

```{r}

cd_hex_pov

legis_path <- here('posts', 'Cartogram', 'data', 'original', 'legislators-current.csv')

legis_tb <- read_csv(legis_path,
                     col_types = cols(.default = col_character()))

congress_tb <- legis_tb %>% 
  filter(!is.na(district)) %>% 
  mutate(district = ifelse(district == '0', '1', district)) 

congress_tb %>% 
  filter(state == "AK")

```


```{r}

c_results_path <- here('posts', 'Cartogram', 'data', 'copy', 'election_results_2024.xlsx')

c_results_tb <- readxl::read_excel(c_results_path) %>% 
  select(State, District, Margin)

```


```{r}

p_results_path <- here('posts', 'Cartogram', 'data', 'copy', '2024 Pres by CD.xlsx')

p_results_tb <- readxl::read_excel(p_results_path) %>% 
  mutate(District = as.character(District)) %>% 
  select(Winner, `Trump Margin %`, State, District)

```


```{r}

enroll_effect_path <- here('posts', 'Cartogram', 'data', 'original', 'data-s4Nmd.csv')

enroll_effect_tb <- read_csv(enroll_effect_path)

toPercentage <- function(str) {
  remove_pct <- sub("%", "", str)
  pctNumber <- as.numeric(remove_pct)
  
  pctNumber / 100
}

enroll_effect_tb2 <- 
  enroll_effect_tb %>% 
    mutate(enrollment_chg_pct = 
             toPercentage(`Percent Change in Enrollment Under Per Capita Cap`),
           expansion_chg_pct =
             toPercentage(`Percent Change in Enrollment Under Per Capita Cap and Elimination of ACA Expansion Match Rate` )) %>% 
    select(State, enrollment_chg_pct, expansion_chg_pct)

```


```{r}

cd_hex <-
  cd_hex_pov %>% 
    mutate(CDLABEL = ifelse(CDLABEL == STATEAB, "1", CDLABEL)) %>% 
    left_join(congress_tb,
              by = c("CDLABEL" = "district",
                     "state"),
              suffix = c('_cd', '_l')) %>% 
    left_join(c_results_tb,
              by = c("CDLABEL" = "District",
                     "state" = "State")) %>% 
    left_join(p_results_tb,
              by = c("CDLABEL" = "District",
                     "state" = "State")) %>% 
  left_join(enroll_effect_tb2,
            by = c("STATENAME" = "State")) %>%
  mutate(enrollment_chg = estimated_medicaid * enrollment_chg_pct,
         expansion_chg = estimated_medicaid * expansion_chg_pct)

analysis <-
  cd_hex %>% 
    st_drop_geometry() %>% 
    tibble()
```

```{r}
analysis %>% 
  select(GEOID, STATEAB, STATENAME, CDLABEL, NAME, population, poverty_n, 
         State_Expanded_Medicaid, Total_Medicaid_Enrollment, estimated_medicaid,
         full_name, party, Margin, Winner, `Trump Margin %`, enrollment_chg_pct,
         expansion_chg_pct)  %>% 
  view()



```


```{r}
cd_hex_geo <- geojson_json(cd_hex) 

cd_hex_out_path <- here('posts', 'Cartogram', 'data', 'copy', 'cd_hex_geo_20250303_4.json')

write_file(cd_hex_geo, cd_hex_out_path)

```




states
```{r}
st_hex_path = here('posts', 'Cartogram', 'data', 'original', 'HexStv30', 'HexStv30.shp')

st_hex <-
  sf::read_sf(st_hex_path)

st_hex_geo <- geojson_json(st_hex) 


st_hex_path_out = here('posts', 'Cartogram', 'data', 'copy', 'st_hex_geo.json')


write_file(st_hex_geo, st_hex_path_out)

```




```{r}
# cd_hex_pov %>% 
#   pull(estimate) %>%# log() %>% 
#   hist()

g <- cd_hex_pov$estimate
m <- mean(g)
std <- sqrt(var(g))

hist(g, density=20, breaks=20, prob=TRUE, 
     xlab="x-variable", ylim=c(0, 2), 
     main="normal curve over histogram")

curve(dnorm(x, mean=m, sd=std), 
      col="darkblue", lwd=2, add=TRUE, yaxt="n")
```

```{r}
g <- cd_hex_pov$estimate %>% log()
m <- mean(g)
std <- sqrt(var(g))

hist(g, density=20, breaks=20, prob=TRUE, 
     xlab="x-variable", ylim=c(0, 2), 
     main="normal curve over histogram")

curve(dnorm(x, mean=m, sd=std), 
      col="darkblue", lwd=2, add=TRUE, yaxt="n")
```


## Observable
https://intro2r.library.duke.edu/interactive.html

This does not work in RStudio, which is great for testing.

Also, quarto preview cannot run on a single file in this project directory structure.

Issue with path in separate drive

OJS code chunk required for ojs_define function

```{r}
#| eval: true

penguin_path <- here('posts', 'Cartogram', 'data', 'original', 'palmer-penguins.csv')

penguins_csv <- read_csv(penguin_path)

ojs_define(data = penguins_csv)

```


```{ojs}
//| eval: true


// needs to be a string literal
//data = FileAttachment(penguin_path).csv({ typed: true })

viewof bill_length_min = Inputs.range(
  [32, 50],
  {value: 35, step: 1, label: "Bill length (min):"}
)
viewof islands = Inputs.checkbox(
  ["Torgersen", "Biscoe", "Dream"],
  { value: ["Torgersen", "Biscoe"],
    label: "Islands:"
  }
)
```

```{ojs}
//| eval: true

filtered = transpose(data).filter(function(penguin) {
  return bill_length_min < penguin.bill_length_mm &&
         islands.includes(penguin.island);
})
```

```{ojs}
//| eval: true

Plot.rectY(filtered, 
  Plot.binX(
    {y: "count"}, 
    {x: "body_mass_g", fill: "species", thresholds: 20}
  ))
  .plot({
    facet: {
      data: filtered,
      x: "sex",
      y: "species",
      marginRight: 80
    },
    marks: [
      Plot.frame(),
    ]
  }
)
```

# Shape file

https://docs.google.com/spreadsheets/d/13XkF59JKzvw4SeSq5mbgIFrJfYjK4amg9JoQE5e9grQ/edit?gid=1250379179#gid=1250379179

