---
title: "Medicaid Cuts Out X"
description: ""
author: "Michael Kuhn"
date: "3/1/2025"
draft: true
output: html_document
format:
  html:
    embed-resources: true
categories:
  - medicaid
  - budget
  - policy
editor_options: 
  chunk_output_type: console
---


```{r}
#| include: false

library(tidyverse)
library(modelr)
library(modelrlibrary(tidycensus)
library(tigris)
library(skimr)
library(dotenv)
library(geojsonio)
library(here)
load_dot_env(file = here::here(".env"))

census_api_key(Sys.getenv('CENSUS_KEY'))

options(tigris_use_cache = TRUE)

```


# Background

Trump budget threatens medicaid funding.

Medicaid is the health insurance program for people with limited resources.



## Medicaid Funding

Federal funding operates through a match at a rate, the federal medical assistance percentage or "FMAP",  that is as low as 50% for states with higher per capita income and as high as 77% for states with lower per capita income.

Additionally, ACA Medicaid expansion states receive a 90% FMAP for qualifying recipients.

The consequence for a state's budget depend not only the number of enrollees in a state, but also on the FMAP rate, which again is higher for low income states.

https://www.kff.org/medicaid/state-indicator/medicaid-expenditures-as-a-percent-of-total-state-expenditures-by-fund/?currentTimeframe=0&sortModel=%7B%22colId%22:%22Total%22,%22sort%22:%22desc%22%7D

figure 4 - budget impact if maintains



## State Options

If medicaid is cut, states may reduce the Medicaid eligibility or increase their own funding to maintain current enrollment levels.

This post will extend the analysis from Kaiser Family Foundation to the congressional district level.



For more detail on medicaid funding at the state level, I recommend the following articles from KFF (formerly the Kaiser Family Foundation).

https://www.kff.org/medicaid/issue-brief/a-medicaid-per-capita-cap-state-by-state-estimates/

https://www.kff.org/medicaid/issue-brief/medicaid-financing-the-basics/

# District Level Mapping

There is not publicly available data for medicaid enrollment at the congressional district level. 

Effects will vary based on the policy details if and when implemented. state specific medicaid eligibility criteria and medicaid expansion enrollment.  




It is X percent of the federal budget. 

Income from census

walk down enrollment state level

https://data.medicaid.gov/dataset/6165f45b-ca93-5bb5-9d06-db29c692a360#data-table



https://www.realclearpolitics.com/elections/live_results/2024/house/

https://guides.loc.gov/election-statistics/presidential-elections

https://centerforpolitics.org/crystalball/the-2024-crossover-house-seats-overall-number-remains-low-with-few-harris-district-republicans/

## Medicaid Enrollment 



```{r}

enrollment_tb <-
  read_csv(here('posts/MedicaidFundingImpact/data/original/pi-dataset-january-2025-release.csv')) %>% 
  rename_with(~ gsub(" ", "_", .x, fixed = TRUE))

enrollment_tb %>% glimpse()

```


```{r}

enrollment_tb %>% 
  select(State_Name, Reporting_Period, State_Expanded_Medicaid, Final_Report,
         Total_Medicaid_Enrollment) %>% 
  arrange(Reporting_Period %>% desc)

```

```{r}

enrollment_total <-
  enrollment_tb %>% 
    select(State_Name, State_Abbreviation, 
           Reporting_Period, State_Expanded_Medicaid, Final_Report,
           Total_Medicaid_Enrollment) %>% 
    filter(Final_Report == "Y") %>% 
    filter(Reporting_Period == max(Reporting_Period))



lm(Total_Medicaid_Enrollment ~ State_Expanded_Medicaid, 
   data = enrollment_total) %>% summary()

```


## Census Poverty Data


Variables

```{r}

v2023 <- load_variables(2023, "acs5/subject", cache = TRUE)

v2023 %>% glimpse()

v2023 %>% 
  filter(name == 'S1701_C01_001') %>% 
  glimpse()

data("fips_codes")

states <- fips_codes %>% 
  select(state, state_code) %>% 
  distinct()

```



```{r}

poverty_cd_s1701 <- get_acs(geography = "congressional district", 
        table = 'S1701',
        year = 2023,
        geometry = TRUE,
        cache_table = TRUE)

poverty_cd_s1701 %>% 
  glimpse()


```

```{r}
# if API fails:

# "E:\Projects\Personal\DataBlog\posts\MedicaidFundingImpact\data\original\productDownload_2025-02-28T164204\ACSST5Y2023.S1702-Data.csv"
# 
# poverty_cd_s1701 <-
#   read_csv(here('posts/MedicaidFundingImpact/data/original/productDownload_2025-02-28T164204/ACSST5Y2023.S1701-Data.csv'),
#            na = c("", "NULL"),
#            skip = 1)
# 
# poverty_cd_s1701 %>% glimpse()
# 
# poverty_cd_s1701 %>% 
#   select(GEO_ID, NAME, S1701_C01_001E, S1701_C02_001E, S1701_C03_001E)

```



```{r}
poverty_cd_s1701_labels <- poverty_cd_s1701 %>% 
  left_join(v2023,
            by = c("variable" = "name")) %>% 
  mutate(state_code = substr(GEOID, 1, 2)) %>% 
  left_join(states)

poverty_cd_s1701_labels %>% 
  glimpse()
```

```{r}

v2023 %>% 
  filter(name == 'S1701_C03_001') %>% 
  glimpse()


poverty_cd_s1701_labels %>% 
  filter(variable == 'S1701_C03_001',
         state != 'PR') %>% 
  shift_geometry() %>% 
  ggplot(aes(fill = estimate)) + 
  geom_sf() + 
  scale_fill_distiller(palette = "RdPu", 
                       direction = 1) + 
  labs(title = "Poverty Rate By Congressional District",
       caption = "Data source: 2023 5-year ACS, US Census Bureau",
       fill = "ACS estimate") + 
  theme_void()

poverty_cd_s1701_labels %>% 
  filter(variable == 'S1701_C02_001',
         state != 'PR') %>% 
  shift_geometry() %>% 
  ggplot(aes(fill = estimate)) + 
  geom_sf() + 
  scale_fill_distiller(palette = "RdPu", 
                       direction = 1) + 
  labs(title = "Poverty Population By Congressional District",
       caption = "Data source: 2023 5-year ACS, US Census Bureau",
       fill = "ACS estimate") + 
  theme_void()
```


```{r}
cd_noPR_geom <-
  poverty_cd_s1701_labels %>% 
  filter(variable == 'S1701_C01_001',
         state != 'PR') %>% 
  select(GEOID, state_code, state, geometry)

pov_data <-
  poverty_cd_s1701_labels  %>% 
  filter(variable %in% c('S1701_C01_001', 'S1701_C02_001'),
         state != 'PR') %>% 
    sf::st_drop_geometry() %>% 
    tibble()


```



We want to confirm relationship. But then, we can apply state-level rates to CDs.

```{r}

cd_pov_data <-
  pov_data %>% 
    filter(state != 'PR') %>% # PR is an outlier
    select(GEOID, NAME, variable, estimate, state_code, state) %>% 
    pivot_wider(names_from = variable,
                 values_from = estimate) %>% 
      rename(population = S1701_C01_001,
             poverty_n = S1701_C02_001)



cd_totals <- # 436, includes DC
  cd_pov_data %>% 
      filter(population > 0) %>% 
      left_join(enrollment_total,
                by = c("state" = "State_Abbreviation")) 


```



```{r}

state_pov_data <-
  pov_data %>% 
    filter(state != 'PR') %>% # PR is an outlier
    group_by(state, variable) %>% 
    summarize(total = sum(estimate, na.rm = T)) %>% 
    ungroup() %>% 
    pivot_wider(names_from = variable,
               values_from = total) %>% 
    rename(population = S1701_C01_001,
           poverty_n = S1701_C02_001)

state_totals <-
  state_pov_data %>% 
    left_join(enrollment_total,
              by = c("state" = "State_Abbreviation"))


state_totals %>% skim()


```

## Enrollment Model



```{r}

state_totals_c <-
  state_totals %>% 
    mutate(enrollment_c = Total_Medicaid_Enrollment - mean(Total_Medicaid_Enrollment),
           population_c = population- mean(population),
           poverty_c = poverty_n - mean(poverty_n))

state_totals_c %>% head()

state_totals_c %>% skim()

enrollment_model_c <-
  lm(enrollment_c ~ State_Expanded_Medicaid * population_c + 
       poverty_c * population_c, 
     data = state_totals_c) 

enrollment_model_c %>% summary()

```

This model accounts for 98% of variance in medicaid enrollment. While there is more to medicaid enrollment than just poverty, a more complicated model would not change this analysis significantly.

Medicaid expansion and population are included in an interaction because many smaller conservative states did not expand medicaid.The interaction accounts for this correlation.

Population and poverty are included in an interaction to account for high population states with high poverty, and vice versa. // is that why?

`intercept` - Medicaid enrollment for states with the average population (`r mean(state_totals$population)`), average number of people in poverty (`r mean(state_totals$poverty_n)`), and that have not expanded Medicaid is -516,400. This is a hypothetical reference point, and its absolute value is not as important as the relative effects of the predictors. This effect is statistically significant at the 0.001 level.

`State_Expanded_MedicaidY` - On average, states that expanded Medicaid have 593,700 more enrollees than non-expansion states, when population and poverty levels are at their means. This effect is statistically significant at the 0.001 level.

`population_c` - The effect of a unit increase in population over average, in a state that has not expanded Medicaid and has an average number of people in poverty, is 0.012 fewer medicaid enrollees. This effect is not statistically significant.

`poverty_c` - The effect of a unit increase in poverty over average, in a state with average population, is 0.82 more medicaid enrollees. This effect is statistically significant at the 0.05 level.

`State_Expanded_MedicaidY:population_c` - in expansion states with average poverty levels, a unit increase in population over average increases the effect of expansion on enrollment by 0.17. This effect is statistically significant at the 0.001 level.

`population_c:poverty_c` - In states that have not expanded medicaid, a unit increase in population increases the effect of poverty by 1.9e-08. This effect is statistically significant at the 0.001 level.

```{r}
enrollment_model <-
  lm(Total_Medicaid_Enrollment ~ State_Expanded_Medicaid * population + 
       poverty_n * population, 
     data = state_totals) 
```

```{r}
# Compute district-level Medicaid estimates
district_data <- cd_totals %>%
  group_by(state) %>%
  mutate(
    weight = (poverty_n * population) / sum(poverty_n * population), # CD poverty share 
    estimated_medicaid = weight * Total_Medicaid_Enrollment
  ) %>%
  ungroup()

# View results
head(district_data)
```

```{r}
district_data

district_data %>% 
  glimpse()

# some districts have 0 population, there are not confirmed districts whatever that means
district_data %>% 
  skim()


cd_state_diff <- 
  district_data %>% 
  group_by(state) %>% 
  summarize(state_enrollment_actual = max(Total_Medicaid_Enrollment),
            state_enrollment_pred = sum(estimated_medicaid)) %>% 
  mutate(difference = state_enrollment_pred - state_enrollment_actual)

cd_state_diff %>% 
  summary()


cd_state_diff  %>% 
  arrange(abs(difference) %>% desc)

```

I think this is good...

```{r}
cd_enrollment_geom <-
  cd_noPR_geom %>% 
    left_join(district_data,
              by = "GEOID")
  
cd_enrollment_geom %>% 
  shift_geometry() %>% 
  ggplot(aes(fill = estimated_medicaid / population)) + 
  geom_sf() + 
  scale_fill_distiller(palette = "RdPu", 
                       direction = 1) + 
  labs(title = "Medicaid Enrollment By Congressional District",
       caption = "Data source: 2023 5-year ACS, US Census Bureau",
       fill = "Estimate") + 
  theme_void()
```




I think this is okay, because we're not concerned about overfitting here.

```{r}
cd_enrollment <-
  cd_state_totals %>% 
    ungroup() %>% 
    add_predictions(enrollment_model5, "enrollment") 


cd_enrollment %>% 
  glimpse()

# some districts have 0 population, there are not confirmed districts whatever that means
cd_enrollment %>% 
  skim()


cd_state_diff <- 
  cd_enrollment %>% 
  group_by(state) %>% 
  summarize(state_enrollment_actual = max(Total_Medicaid_Enrollment),
            state_enrollment_pred = sum(enrollment * population)) %>% 
  mutate(difference = state_enrollment_pred - state_enrollment_actual)

cd_state_diff %>% 
  summary()


cd_state_diff  %>% 
  arrange(abs(difference) %>% desc)

```


```{r}
cd_enrollment_geom <-
  cd_noPR_geom %>% 
    left_join(cd_enrollment,
              by = "GEOID")
  
cd_enrollment_geom %>% 
  shift_geometry() %>% 
  ggplot(aes(fill = enrollment * population)) + 
  geom_sf() + 
  scale_fill_distiller(palette = "RdPu", 
                       direction = 1) + 
  labs(title = "Medicaid Enrollment By Congressional District",
       caption = "Data source: 2023 5-year ACS, US Census Bureau",
       fill = "Estimate") + 
  theme_void()
```

## GeoJSON

```{r}


cd_hex_path = here('posts', 'MedicaidFundingImpact', 'data', 'original', 'HexCDv31', 'HexCDv31.shp')

cd_hex <-
  sf::read_sf(cd_hex_path)

cd_hex_pov <-
  cd_hex %>% 
    left_join(district_data,
              by = 'GEOID')


cd_hex_geo <- geojson_json(cd_hex_pov) 

geo_path <- here('posts', 'MedicaidFundingImpact', 'data', 'copy', 'cd_hex_geo_enrollment.json')

write_file(cd_hex_geo, geo_path)

```




